Recursive partitioning of linear growth curve models

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
make notation of b_i more consistent
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3245 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 304253a3c76033f9e2dc86fcbf6716e3ce23b7aa

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which observation~$i$ falls into. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).

Thus, when it is known which node~$j$ each observation~$i$ falls into, the GLMM specified by Equations~\ref{eq:expected_value}--\ref{eq:GLMMtree} can be fitted ``as usual'', yielding local subgroup-specific fixed-effect estimates $\hat \beta_j$ and global random-effect estimates $\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm where the initialization simply assumes zero random effects.

\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}

@@

Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$.

>>>>

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which subject~$i$ falls into. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b_i$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma$.

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML). Thus, when it is known which node~$j$ each subject~$i$ falls into, the GLMM specified by Equation~\ref{eq:GLMMtree} can be fitted ``as usual'', yielding \emph{local} subgroup-specific fixed-effect estimates $\hat \beta_j$ and \emph{global} random-effect estimates $\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm where the initialization simply assumes zero random effects.

\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting $r$ and all values $\hat{b}_{i,(r)}$ to 0.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b}_{i,(r-1)}$ as an offset.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{i,(r)}$ from the estimated model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}

@@

Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$ from the previous step.
